Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-25 appendix.tex
put eqns in table and change some grammar

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
\]
Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

>>>>

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$ is the unique maximum of the divergence term for $x_1$ and $x_2$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$ with $w \neq 0$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,2}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,3}(x_2)\\
&\qquad +\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,4}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,5}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
We have the second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
As $x_1,y_1,x_2,y_2\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood is negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau_1, \tilde{t})$ & $\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,\tilde{t})$\\
    \hline
     $\{1\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2)\\
                    (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2)\\
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
                \end{array}$\\
    \hline
     $\{1,2\}$    &
    $\begin{array}{l}
                    \emptyset\\
                    \{1\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}$&
    $\begin{array}{l}
                    (1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2)\\
                    (1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2)\\
                    (1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2)\\
                    (1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2)
                \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for site splits $\{1\}$ and $\{1,2\}$ and ancestral state partitions $\ancestralSplitPartition_j$ on the InvFels tree $\tau_1$ with general branch parameters $\tilde{t}$.
All values multiplied by $1/32$.}
\label{tab:edge-cases}
\end{table}

We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
Site split $\emptyset$ always attains a maximum at ancestral state split $\emptyset$ provided at least one branch fidelity is not exactly zero.
Table~\ref{tab:edge-cases} gives the possible forms for the partial likelihood for two cases of site splits.
For site split $\{1\}$ we ignore ancestral state split $\{2\}$ as its partial likelihood is always less than the ancestral state split $\emptyset$ except for the case $y_1=y_2=w=0$.
Comparing the remaining ancestral state splits reveals that there is an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
We reach the same conclusions doing these comparisons for site splits $\{2\}$, $\{3\}$, and $\{1,2,3\}$.

The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
For site split $\{1,2\}$, we ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization occurs when
\[
(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2),
\]
or, in other words, when the ancestral state splits $\emptyset$ and $\{1,2\}$ both maximize the partial likelihood.
This can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$, though, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
To resolve this ambiguity, we look to the divergence quanitity that is maximized when $x_1=x_2$ and $y_1=y_2$, and, thus, the full joint likelihood obtains a unique maximum when $x_1=x_2$ and $y_1=y_2$.

Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
